---
bibliography: ./bibtex/library.bib # Path to the bibtex file.
fontsize: 12pt
linestretch: 1.3 
geometry: top=1.75in,left=1in,right=1in,bottom=1.75in # Backed out from LaTeX PDF output.
header-includes:
- \usepackage{url}
- \usepackage{float}
- \usepackage{amssymb}
- \usepackage{amsmath}
- \usepackage{makeidx}
- \usepackage{fancyhdr}
- \usepackage{verbatim}
- \usepackage{pdfpages}
- \usepackage{epstopdf}
- \usepackage{hyperref}
- \usepackage{pdflscape}
- \usepackage[all]{hypcap}
- \usepackage[pdftex]{thumbpdf}
- \usepackage{./styles/texnames}
- \usepackage[number=none]{./styles/glossary}
linkcolor: blue
urlcolor: blue
documentclass: report
classoption: a4paper,oneside
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    keep_tex: false # Change this to true if you want the output compiled .tex (useful for debugging).
  html_document: default # Compiling to HTML does not work properly (yet).
---


```{r global_options, include=FALSE}
# This knitr code is required to compile to PDF.
knitr::opts_chunk$set(fig.path = 'figures/',
                      fig.pos = 'h',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```


```{r setup, include = TRUE, cache = FALSE}
# RefManageR allows you to manage citations and references from your bibtex. 
# Currently the inline numeric links are not working, they link to the top of the document?
library(RefManageR)
bib <- ReadBib("./bibtex/library.bib")
BibOptions(check.entries = FALSE, 
           style = "markdown", 
           bib.style = "numeric", 
           cite.style = "numeric")
```


<!-- LATEX taken from Will's MSc Dissertation template. 
Link: http://cirg.cs.up.ac.za/visitPage.php?pageID=17060 -->
\hypersetup{
	pdfhighlight=/N,
	pdffitwindow=true,
	pdfstartview=Fit,
	plainpages=false,
	breaklinks=true,
	colorlinks=true,
	pageanchor=false,
	linkcolor=blue,
	citecolor=blue,
	filecolor=blue,
	menucolor=blue,
	pagecolor=blue,
	urlcolor=blue,
}


\hypersetup{
	pdftitle    = {Title},
	pdfauthor   = {Name Surname <name@gmail.com>},
	pdfsubject  = {Subject},
	pdfkeywords = {Something, Some other thing},
}


\pdfadjustspacing=1


\setlength{\textwidth}{158mm}
\setlength{\headheight}{15pt}
\setlength{\unitlength}{1pt}
\setlength{\footnotesep}{5mm}


\newlength{\originalbaselineskip}
\setlength{\originalbaselineskip}{\baselineskip} \linespread{1.3}


\newcounter{listcount}


\renewcommand{\glossaryname}{Acronyms}
\renewcommand{\gloskip}{}
\setlength{\namewidth}{79pt}


\newcommand{\idxbf}[1]{\textbf{\hyperpage{#1}}}


\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.0}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.9}


\newfloat{graph}{tbp}{lgf}
\floatname{graph}{Graph}


\newfloat{algorithm}{tbp}{loa}
\floatname{algorithm}{Algorithm}


\setcounter{topnumber}{1}
\setcounter{bottomnumber}{0}


\newlength{\abovesubfiglabelskip}
\setlength{\abovesubfiglabelskip}{0.5\abovecaptionskip}


\newacronym{ANN}{Artificial Neural Network}{name=ANN, description=Artificial Neural Network}


\index{ANN|see{Artificial Neural Network}}


<!-- Include the cover page which includes abstract and stuff. -->
```{r child = './front.Rmd'}
```


\hypersetup{pageanchor=true}
\pagestyle{fancy}
\fancypagestyle{headings}{
	\fancyhead[RO,LE]{\thepage}
	\fancyhead[LO]{\sf\nouppercase{\leftmark}}
	\fancyhead[RE]{\sf\nouppercase{\rightmark}}
	\fancyfoot{}
	
	\renewcommand{\headrulewidth}{0.4pt}
}


<!-- Include chapter 0 a.k.a the introduction -->
```{r child = './chapters/chapter0.Rmd'}
```


<!-- Include chapter one. -->
```{r child = './chapters/chapter1.Rmd'}
```


<!-- Include chapter two. -->
```{r child = './chapters/chapter2.Rmd'}
```


<!-- Include chapter two. -->
```{r child = './chapters/chapter3.Rmd'}
```


<!-- Include chapter two. -->
```{r child = './chapters/chapter4.Rmd'}
```


\cleardoublepage
\phantomsection


\chapter*{Bibliography}\label{bibliography}
\addcontentsline{toc}{chapter}{Bibliography}
\bibliographystyle{plain}
\bibliography{dissertation}


```{r results = "asis", echo = FALSE}
PrintBibliography(bib, .opts = list(check.entries = FALSE, sorting = "ynt"))
```
